---
title: "R Notebook"
output: html_notebook
---

This is a data cleaning notebook. 

About Dataset

The Paycheck Protection Program (PPP) is a $953-billion business loan program established by the United States federal government, led by the Donald Trump administration in 2020 through the Coronavirus Aid, Relief, and Economic Security Act (CARES Act) to help certain businesses, self-employed workers, sole proprietors, certain non-profit organizations, and tribal businesses continue paying their workers.

The Paycheck Protection Program allows entities to apply for low-interest private loans to pay for their payroll and certain other costs. The amount of a PPP loan is approximately equal to 2.5 times the applicant's average monthly payroll costs. In some cases, an applicant may receive a second draw typically equal to the first. The loan proceeds may be used to cover payroll costs, rent, interest, and utilities. The loan may be partially or fully forgiven if the business keeps its employee counts and employee wages stable. The program is implemented by the U.S. Small Business Administration. The deadline to apply for a PPP loan was March 31, 2021.
Some economists have found that the PPP did not save as many jobs as purported and aided too many businesses that were not at risk of going under. They noted that other programs, such as unemployment insurance, food assistance, and aid to state and local governments, would have been more efficient at strengthening the economy. Opponents to this view note that the PPP functioned well to prevent business closures and cannot be measured on the number of jobs saved alone.

According to a 2022 study, the PPP: cumulatively preserved between 2 and 3 million job-years of employment over 14 months at a cost of USD 169K to USD 258K per job-year retained. These numbers imply that only 23 to 34 percent of PPP dollars went directly to workers who would otherwise have lost jobs; the balance flowed to business owners and shareholders, including creditors and suppliers of PPP-receiving firms. Program incidence was ultimately highly regressive, with about three-quarters of PPP funds accruing to the top quintile of households. PPP's breakneck scale-up, its high cost per job saved, and its regressive incidence have a common origin: PPP was essentially not specifically targeted because the United States lacked the administrative infrastructure to do otherwise. Harnessing modern administrative systems, other high-income countries were able to better target pandemic business aid to firms in financial distress. Building similar capacity in the U.S. would enable improved targeting when the next pandemic or other large-scale economic emergency inevitably arises.

Additional Information:
Created: April 5, 2022
Format: CSV
License: Other (Public Domain)
Size: 428.6 MB

Load appropriate libraries and dataset
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(formattable)
library(gt)

#Load Data
ppp_loans <- readr::read_csv("./ppp_loans.csv")
dim(ppp_loans)
str(ppp_loans)
```

There are columns that are not needed, these fields have limited analytical value.

```{r}
#delete columns that are not needed for analysis
ppp_loans = ppp_loans[-c(3,4,6,7,9,13,17,18,19,20,
               21,22,23,31,36,37,39,40,41,42,43,45,47,48)]

dim(ppp_loans)
str(ppp_loans)
```

Handling dates
```{r}
#Convert Date Approved to Date datatype
ppp_loans$DateApproved <- as.Date(ppp_loans$DateApproved, "%m/%d/%Y")
class(ppp_loans$DateApproved)
unique(ppp_loans$DateApproved)
min(ppp_loans$DateApproved)
max(ppp_loans$DateApproved)

#Add Year-Month to trend data
ppp_loans$Month_Yr = strftime(ppp_loans$DateApproved,format="%Y-%m")
unique(ppp_loans$Month_Yr)
```
Example NA and handle null values where appropriate
```{r}
#examine na across columns
map(ppp_loans, ~sum(is.na(.)))

sum(is.na(ppp_loans$BorrowerName))
ppp_loans[is.na(ppp_loans$BorrowerName),]

sum(is.na(ppp_loans$BorrowerState))

ppp_loans = ppp_loans %>%
  mutate(BorrowerState = if_else(is.na(BorrowerState) , ProjectState, BorrowerState))

sum(is.na(ppp_loans$BorrowerState))

ppp_loans[is.na(ppp_loans$BorrowerState),]

ppp_loans = ppp_loans %>%
  mutate(BusinessType = if_else(is.na(BusinessType) , "Unknown", BusinessType))

sum(is.na(ppp_loans$BusinessType))

ppp_loans = ppp_loans %>%
  mutate(NonProfit = if_else(is.na(NonProfit) , "N", NonProfit))

sum(is.na(ppp_loans$NonProfit))

ppp_loans[is.na(ppp_loans$BusinessAgeDescription),]

ppp_loans = ppp_loans %>%
  mutate(BusinessAgeDescription = if_else(is.na(BusinessAgeDescription) , 
                                          "Unanswered", BusinessAgeDescription))

sum(is.na(ppp_loans$BusinessAgeDescription))

#Delete columns: ProjectCity, ProjectCountyName, ProjectState CD
ppp_loans = subset(ppp_loans, select = -c(ProjectCity,
                                          ProjectCountyName,ProjectState,CD))
dim(ppp_loans)
tail(ppp_loans)

#examine na across columns
map(ppp_loans, ~sum(is.na(.)))

ppp_loans = ppp_loans %>% 
  mutate(UndisbursedAmount = ifelse(is.na(UndisbursedAmount), 
                                    0, UndisbursedAmount))

ppp_loans = ppp_loans %>% 
  mutate(ForgivenessAmount = ifelse(is.na(ForgivenessAmount), 
                                    0, ForgivenessAmount))

ppp_loans = ppp_loans %>% 
  mutate(JobsReported = ifelse(is.na(JobsReported), 
                               0, JobsReported))

```
The North American Industry Classification System (NAICS) is the standard used by Federal statistical agencies in classifying business establishments for the purpose of collecting, analyzing, and publishing statistical data related to the U.S. business economy. NACIS codes are important for industry analysis, we will segment the loans by Industry to better understand which industries were hardest hit by COVID

```{r}
#add 2 digit nacis code and industry mapping
nacis <- readr::read_csv("./two_digit_naics_code.csv")

#extract 2 digit nacis codes from ppp loans data
ppp_loans = ppp_loans%>%
  mutate(NAICSCode2 = as.numeric(gsub("(^\\d{2}).*", "\\1", ppp_loans$NAICSCode)))

sort(unique(ppp_loans$NAICSCode2))

unique(nacis$NAICSCode2)

# using join to add Industry dimension 
ppp_loans = ppp_loans%>%left_join(nacis)

ppp_loans = ppp_loans %>% 
  mutate(Industry = ifelse(is.na(Industry), "Unknown", Industry))

unique(ppp_loans$Industry)
map(ppp_loans, ~sum(is.na(.)))
```

Remove metrics that are not needed for the analysis, we will focus on Current Approval Amount

```{r}
ppp_loans%>%
  group_by(BusinessType)%>%
  summarise(init_amt = sum(InitialApprovalAmount),
            curr_amt = sum(CurrentApprovalAmount),
            undis_amt = sum(UndisbursedAmount),
            forg_amt = sum(ForgivenessAmount)) 
  
#The UndisbursedAmount is small and we will ignore it for this analysis
ppp_loans = subset(ppp_loans, select = -c(UndisbursedAmount))
```

Add State and Population to data. We are adding an estimate of small businesses in the state based on 2017 estimates and payroll in USD. The dataset also has politics (red vs. blue) based on 2016 Presidential Election results.

```{r}
state_rb <- readr::read_csv("./red_vs_blue_state.csv")
state_rb = state_rb%>%mutate(BorrowerState = str_trim(state_rb$BorrowerState))
state_rb = state_rb%>%mutate(politics = ifelse(is.na(Red), "Blue", "Red"))

ppp_loans = ppp_loans%>%left_join(state_rb, by="BorrowerState")
dim(ppp_loans)

state_pop <- readr::read_csv("./State_POP_SB_Payroll.csv")
state_pop = state_pop%>%mutate(BorrowerState = 
                                 str_trim(state_pop$BorrowerState))
ppp_loans = ppp_loans%>%left_join(state_pop)
dim(ppp_loans)

ppp_loans%>%
  select(Population,
         Small_Business_Cnt,
         Employment_Estimate_2017,
         Annual_payroll_in_thousand)
#check na once more
map(ppp_loans, ~sum(is.na(.)))
```

